Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2018-01-25 cabios-template2/bigsnpr-review.tex
new draft of review

\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge severely slowing down genomic analyses. Specialized software for every part of the analysis pipeline have been developed to handle large genomic data. However, combining all these software into a single data analysis pipeline might be technically difficult. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} \& \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Genome-wide datasets produced for association studies have dramatically increased in size over the past years. A range of software and data formats have been developed to perform essential pre-processing steps and data analysis, often optimizing each of these steps within a dedicated implementation. This diverse and extremely rich software environment has been of tremendous benefit for the genetic community. However, it has two limitations: analysis pipelines are becoming very complex and researchers have limited access to diverse analysis tools due to growing data sizes.

Consider first the basic tools necessary to perform a standard genome-wide analysis. Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK, available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}. Similarly, quality control software for genome-wide analysis have been developed such as PLINK and the Bioconductor package GWASTools \cite[]{Gogarten2012}.
There are also several software for the computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, including EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}.
Then, implementation of GWAS analyses also depends on the data format and model analyzed. For example, the software ProbABEL \cite[]{Aulchenko2010} or SNPTEST \cite[]{Marchini2010} can handle dosage data, 
{\color{red}
while PLINK version 1.9 is limited to best guess genotypes because of its input file format. Note that PLINK 2.0 will completely handle this type of data, but is still in the alpha stage of development. 
}
Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred \cite[]{Vilhjalmsson2015} and PRSice \cite[]{Euesden2015}, which provide prediction for quantitative traits or disease risks based on multiple genetic variants. As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions. Overall, this might be a barrier to data exploration. 
%Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.

Secondly, increasing size of genetic datasets is the source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers. This is particularly a burden for commonly used analysis languages such as R, Python and Perl. Solving the memory issues for these languages would give access to a broad range of tools for data analysis that have been already implemented. Hopefully, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.

%\enlargethispage{12pt}

@@

For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions, for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD (Figure~\ref{fig:svd}). While the function based on eigenvalue decomposition is at least quadratic in the smallest dimension, the function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}. Package bigstatsr use the same PCA algorithm as FlashPCA2 called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra. The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the FBM format, which enables parallel computations and easier subsetting.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{svd.pdf}}
\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array, with 3 different methods for thinning SNPs.}\label{fig:svd}
\end{figure}

SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}. There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium, two of them named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}. 
First, pruning, the most naive approach, is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
A variant of pruning is clumping. 
{\color{red}
Clumping is useful if a statistic is available to sort the SNPs by importance. Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome. For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance. Note that this choice does not come out of nowhere because in the pruning algorithm of PLINK, when two nearby SNPs are correlated, only the one with the higher MAF is kept. Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]). So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping . In practice, we tested the effect of using clumping instead of pruning for the thinning step before PCA and the results were very similar.
}

As mentioned above, the third approach that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}. Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in long-range LD regions while computing PCs\footnote{https://goo.gl/8TngVE}. 
However, the pattern of LD might be population specific, so we developed an algorithm that automatically detects these regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}. Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions. Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA. This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}. 
We also implement statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $K$ is the number of principal components and $L$ is the number of other covariates (such as the age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.

The R packages also implement functions to compute Polygenic Risk Scores using two approaches. 
The first method is the widely-used Clumping + Thresholding (C+T, also called ``Pruning + Thresholding'' in the literature) model based on univariate GWAS summary statistics as described in previous equations. Under the C+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value. The SNPs are first clumped (C) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). A polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Chatterjee2013,Dudbridge2013,Golan2014}. 
The second approach does not use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables at once, optimally accounting for correlation between predictors \cite[]{Abraham2012}. The currently available models are very fast sparse linear and logistic regressions. These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Friedman2010,Tibshirani1996,Zou2005}. Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}. The implementation of these algorithms is based on modified versions of functions available in the R package biglasso \cite[]{Zeng2017}. These modifications allow to include covariates in the models, to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr, to remove the need of the choice of the regularization parameter and to fasten the overall algorithms.

@@

All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the Supplementary Data. 
{\color{red}
In supplementary notebook ``public-data'', we provide some open-access data of domestic dogs so that users can test our code and functions on a moderate size dataset with 4342 samples and 145,596 SNPs \cite[]{hayward2016complex}.
}

@@

We present the results for three different analyses. First, we illustrate the application of R packages bigstatsr and bigsnpr. Secondly, we compare the performance of the R packages to the performance obtained with PLINK and FastPCA (EIGENSOFT). Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes {\color{red} (for genotyped SNPs only)}. We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. We compare performance on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores). For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

We performed an association study and computed a polygenic risk score for the Celiac cohort. 
The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}. 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\%. The 0.04\% missing genotype values were imputed with the XGBoost method. If we used a standard R matrix to store the genotypes, this data would require 32GB of memory. On the disk, the ``.bed'' file requires 1GB and the ``.bk'' file (storing the FBM) requires 4GB.

We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of the Celiac genotype matrix and to visualize them (Figure~\ref{fig:pca}). We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while adjusting for PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}). As illustrated in the supplementary data, the whole pipeline is user-friendly and requires only 20 lines of R code.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=200pt]{celiac-pca}}
\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=235pt]{celiac-gwas-cut}}
\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. Some SNPs in chromosome 6 have p-values smaller than the $10^{-30}$ threshold used for vizualisation purposes.}\label{fig:gwas}
\end{figure}

%The Celiac dataset is relatively small as compared to modern genetic cohorts. 
To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs. The GWAS analysis completed in approximately 11 hours using the aforementioned desktop computer. The GWAS analysis was composed of four main steps. 
First we read from PLINK files in our format "bigSNP" in 1 hour.
Then, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs in 5.4h. Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs in 1.8h. Finally, we performed a linear association test on the complete 500K dataset for each of the 1M SNPs, using the 10 first PCs as covariables in 2.9h.

\subsection{Method Comparison}

We first compared the GWAS and PRS computations obtained with the R packages to the ones obtained with PLINK 1.9 and EIGENSOFT 6.1.4.
For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g.\ pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table~\ref{tab:bench-gwas}). Overall, the computations with our two R packages for an association study and a polygenic risk score are of the same order of magnitude as when using PLINK and EIGENSOFT (Tables~\ref{tab:bench-gwas} and~\ref{tab:bench-prs}). However, the whole analysis pipeline makes use of R calls only; there is no need to write temporary files and functions have parameters which enable subsetting of the genotype matrix without having to copy it.

\begin{table}[!tpb]
\begin{center}
\begin{adjustbox}{max width=235pt}
\begin{tabular}{|c|c|c|c|}
\hline
\multirow{3}{*}{Operation~~\textbackslash~~Software} &   \multicolumn{3}{c|}{Execution times (in seconds)} \\
 \cline{2-4}
 & FastPCA & bigstatsr  & SNPRelate \\
 & PLINK 1.9 & bigsnpr & GWASTools \\
\hline
Reading PLINK files & n/a & 6 / 20 & 12 /  \\
Pruning & 4 / 4 & 14 / 52 & 33 / \\  
Computing 10 PCs & 304 / 315 & 61 / 180 &  /  \\
GWAS (binary phenotype) & 336 / 293 & 301 / 861 & 15806 /  \\
GWAS (continuous phenotype) & 1339 /  & 9 /  & 5933 /  \\
\hline
Total & 649 / 612 & 378 / 1113 &  /  \\
\hline
\end{tabular} 
\end{adjustbox}
\end{center}
\caption{Execution times with bigstatsr and bigsnpr compared to PLINK 1.9 and FastPCA (EIGENSOFT) and also to R packages SNPRelate and GWASTools for making a GWAS for the Celiac dataset. The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop (2 cores used and 8GB of RAM).} 
\label{tab:bench-gwas}
\end{table}

\begin{table}[!tpb]
\begin{center}
\begin{adjustbox}{max width=235pt}
\begin{tabular}{|c|c|c|}
\hline
\multirow{2}{*}{Operation~~\textbackslash~~Software} &   \multicolumn{2}{c|}{Execution times (in seconds)} \\
 \cline{2-3}
 & PRSice & bigstatsr and bigsnpr \\
\hline
Reading PLINK files &  & 6 / 20 \\
Reading summary stats &  & 4 /   \\
Clumping &  & 9 /  \\
PRS &  & 2 /  \\
Compute pvalues &  & 1 /  \\
\hline
Total & 22 /  & 22 /  \\
\hline
\end{tabular} 
\end{adjustbox}
\end{center}
\caption{Execution times with bigstatsr and bigsnpr compared to PRSice for making a PRS on the Celiac dataset based on summary statistics for Height. The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop (2 cores used and 8GB of RAM).}
\label{tab:bench-prs}
\end{table}

>>>>

\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. {\color{red} This increase in data size is a major challenge severely slowing down genomic analyses, leading to some software becoming obsolete and researchers having limited access to diverse analysis tools.} \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for the analysis of large scale genomic data to be performed within R. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} \& \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

{\color{red}
Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
As a consequence, it has been required to continuously optimize algorithms and this has led to the reimplementation of some software that would have become obsolete otherwise. For computing Principal Component Analysis, commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to smartpca and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}. 
PLINK 1.07, which is a central tool in the analysis of genotype data, has been replaced by PLINK 1.9, and soon PLINK 2.0 \cite[]{chang2015second,Purcell2007}.

Increasing size of genetic datasets is a source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers.
This is particularly a burden for commonly used analysis languages such as R.
For analyzing genotype datasets in R, a range of software are available, including R packages snpStats, GenABEL, SNPRelate and GWASTools. Packages snpStats and GenABEL, which have been widely used and cited, are now outdated with current data sizes -- package snpStats still uses in-memory data and package GenABEL does not handle standard binary PLINK files and has not been updated since 2013.
Solving the memory issues for languages such as R would give access to a broad range of tools for data analysis that have already been implemented. Hopefully, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
}

@@

For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD (Figure~\ref{fig:svd}). While the function based on eigenvalue decomposition is at least quadratic in the smallest dimension, the function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}. Package bigstatsr uses the same PCA algorithm as FlashPCA2 called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra. The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the FBM format, which enables parallel computations and easier subsetting.

\begin{figure}[!htpb]
\centerline{\includegraphics[width=235pt]{svd.pdf}}
\caption{Functions available in packages bigstatsr and bigsnpr for the computation of a partial Singular Value Decomposition of a genotype array, with 3 different methods for thinning SNPs.}\label{fig:svd}
\end{figure}

SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}. There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium. Two of them, named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}. 
First, pruning is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
{\color{red}
Clumping is useful if a statistic is available for sorting the SNPs by importance. Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome. 
For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the MAF as the statistic of importance. 
Note that this choice does not come out of nowhere because in the pruning algorithm of PLINK, when two nearby SNPs are correlated, only the one with the highest minor allele frequency (MAF) is kept.
Yet, in some worst-case scenario, this pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]). 
So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr. In practice, we tested the effect of using clumping with the MAF instead of pruning for the thinning step before PCA and the results were very similar.
}

As mentioned above, the third approach that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}. Long-range LD regions for the human genome are available as an online table\footnote{https://goo.gl/8TngVE} that package bigsnpr can use to discard SNPs in long-range LD regions begore computing PCs. 
However, the pattern of LD might be population specific, so we developed an iterative algorithm that automatically detects these long-range LD regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping (with MAFs), then outliers SNPs are detected using the robust Mahalanobis distance as implemented in method pcadapt \cite[]{Luu2017}. Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions. Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA. This algorithm is implemented in function snp\_autoSVD of package bigsnpr and will be referred by this name in the rest of the paper.

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcomes \cite[]{Zheng2012}. 
Package bigstatsr implements statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $K$ is the number of principal components and $L$ is the number of other covariates (such as the age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
{\color{red}
These tests can be used to perform genome-wide association studies (GWAS) and are very fast due to the use of ingenious implementations, partly based on previous work by \cite{sikorska2013gwas}. 
}
%For a continuous phenotype $y$ and a SNP $s$ and some covariables $X$ (including a column of 1s for the intercept), we can compute $s^* = s - X(X^TX)^{-1} X^T s$ and $y^* = s - X(X^TX)^{-1} X^T y$ (orthogonal projections relative to the space of covariables).

The R packages also implement functions for computing Polygenic Risk Scores using two methods. 
{\color{red} The first method is the widely-used ``Clumping + Thresholding'' (C+T, also called ``Pruning + Thresholding'' in the literature) model based on univariate GWAS summary statistics as described in previous equations.}
Under the C+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value. The SNPs are first clumped (C) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). A polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Chatterjee2013,Dudbridge2013,Euesden2015}. 
On the contrary, the second approach does not use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables \textit{at once}, optimally accounting for correlation between predictors \cite[]{Abraham2012}. The currently available models are very fast sparse linear and logistic regressions. These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Friedman2010,Tibshirani1996,Zou2005}. Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}. The implementation of these algorithms is based on modified versions of functions available in the R package biglasso \cite[]{Zeng2017}. These modifications allow to include covariates in the models, to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr and to remove the need of the choice of the regularization parameter.

@@

All the code used in this paper along with results, such as execution times and figures, are available as HTML R notebooks in the Supplementary Materials. 
{\color{red}
In supplementary notebook ``public-data'', we provide some open-access data of domestic dogs so that users can test our code and functions on a moderate size dataset with 4342 samples and 145,596 SNPs \cite[]{hayward2016complex}.
}

@@

We present the results for four different analyses. 
First, we illustrate the application of R packages bigstatsr and bigsnpr. 
Secondly, by performing two GWAS, we compare the performance of bigstatsr and bigsnpr to the performance obtained with FastPCA (EIGENSOFT 6.1.4) and PLINK 1.9, and also two R packages SNPRelate and GWASTools \cite[]{chang2015second,Galinsky2016,Gogarten2012,zheng2012high}. For the step concerning the computation of 10 principal components, we further compare methods on larger datasets.
Thirdly, by performing a standard PRS analysis, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (for genotyped SNPs only). 
We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. 
We compare performance on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores). For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs with more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency (MAF) lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}. 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\%. The 0.04\% missing genotype values were imputed with the XGBoost method. If we used a standard R matrix to store the genotypes, this data would require 32GB of memory. On the disk, the ``.bed'' file requires 1GB and the ``.bk'' file (storing the FBM) requires 4GB.

We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of the Celiac genotype matrix and to visualize them (Figure~\ref{fig:pca}). We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while adjusting for PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}). As illustrated in the supplementary data, the whole pipeline is user-friendly, requires only 20 lines of R code and there is no need to write temporary files or objects because functions of packages bigstatsr and bigsnpr have parameters which enable subsetting of the genotype matrix without having to copy it.

\begin{figure}[!htpb]
\centerline{\includegraphics[width=200pt]{celiac-pca}}
\caption{Principal Components of the celiac cohort genotype matrix produced by package bigstatsr.}\label{fig:pca}
\end{figure}

\begin{figure}[!htpb]
\centerline{\includegraphics[width=235pt]{celiac-gwas-cut}}
\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. Some SNPs in chromosome 6 have p-values smaller than the $10^{-30}$ threshold used for vizualisation purposes.}\label{fig:gwas}
\end{figure}

%The Celiac dataset is relatively small as compared to modern genetic cohorts. 
To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs. The GWAS analysis completed in approximately 11 hours using the aforementioned desktop computer. The GWAS analysis was composed of four main steps. 
First we read from PLINK files in our format ``bigSNP'' in 1 hour.
Then, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs in 5.4h. Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs in 1.8h. Finally, we performed a linear association test on the complete 500K dataset for each of the 1M SNPs, using the 10 first PCs as covariables in 2.9h.

\subsection{Performance and precision comparisons}

First, we compared the GWAS computations obtained with bigstatsr and bigsnpr to the ones obtained with PLINK 1.9 and EIGENSOFT 6.1.4, and also two R packages SNPRelate and GWASTools.
For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g.\ pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table~\ref{tab:bench-gwas}). 
Overall, performing a GWAS on a binary outcome with our two R packages, bigstatsr and bigsnpr, is as fast as when using EIGENSOFT and PLINK, and much faster than when using R packages SNPRelate and GWASTools. 
For performing an association study on a continuous outcome, we report a dramatic increase in performance by using bigstatsr and bigsnpr, making it possible to perform such analysis in less than 2 minutes for a relatively large dataset as the Celiac dataset.
Note that the PC scores obtained are more precise (see the last paragraphe of this subsection), which is also the case for the p-values as compared to PLINK (see supplementary notebook ``GWAS-comparison'').

\begin{table}[!htpb]
\begin{center}
\begin{adjustbox}{max width=235pt}
\begin{tabular}{|c|c|c|c|}
\hline
\multirow{3}{*}{Operation~~\textbackslash~~Software} &   \multicolumn{3}{c|}{Execution times (in seconds)} \\
 \cline{2-4}
 & FastPCA & bigstatsr  & SNPRelate \\
 & PLINK 1.9 & bigsnpr & GWASTools \\
\hline
Reading PLINK files         &     n/a     &   6 /  20 &    13 /    33 \\
Pruning                     &    4 /    4 &  14 /  52 &    33 /    32 \\  
Computing 10 PCs            &  305 /  314 &  58 / 183 &   323 /   535 \\
GWAS (binary phenotype)     &  337 /  284 & 291 / 682 & 16220 / 17425 \\
GWAS (continuous phenotype) & 1348 / 1633 &  10 /  23 &  6115 /  7101 \\
\hline
Total (binary)              &  646 /  602 & 369 / 937 & 16589 / 18025 \\
Total (continuous)          & 1657 / 1951 &  88 / 278 &  6484 /  7701 \\
\hline
\end{tabular} 
\end{adjustbox}
\end{center}
\caption{Execution times with bigstatsr and bigsnpr compared to PLINK 1.9 and FastPCA (EIGENSOFT) and also to R packages SNPRelate and GWASTools for making a GWAS for the Celiac dataset (15,155 individuals and 281,122 SNPs). The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop (2 cores used and 8GB of RAM).} 
\label{tab:bench-gwas}
\end{table}

Secondly, we compared the PRS analysis performed with the R packages to the one using PRSice-2. There are 5 main steps in such a analysis (see table~\ref{tab:bench-prs}), including 4 steps handled with functions of package bigstatsr and bigsnpr. The remaining step is the reading of summary statistics which is performed with function fread of R package data.table, a widely-used fast and convenient reading function. [FINIR TIMING]

\begin{table}[!htpb]
\begin{center}
\begin{adjustbox}{max width=235pt}
\begin{tabular}{|c|c|c|}
\hline
\multirow{2}{*}{Operation~~\textbackslash~~Software} &   \multicolumn{2}{c|}{Execution times (in seconds)} \\
 \cline{2-3}
 & PRSice & bigstatsr and bigsnpr \\
\hline
Reading PLINK files   &  & 6 / 20 \\
Reading summary stats &  & 4 / 6  \\
Clumping              &  & 9 / 31 \\
PRS                   &  & 2 / 33? \\
Compute pvalues       &  & 1 / 1  \\
\hline
Total                 & 22 / 29 & 22 /  \\
\hline
\end{tabular} 
\end{adjustbox}
\end{center}
\caption{Execution times with bigstatsr and bigsnpr compared to PRSice for making a PRS on the Celiac dataset based on summary statistics for Height. The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop (2 cores used and 8GB of RAM).}
\label{tab:bench-prs}
\end{table}
